Untitled

### 2022-05-05 Papers/Polymethylation_scores/02_abstract.Rmd
revisions may5

Prenatal maternal smoking is associated with low birthweight, neurological disorders, and asthma in exposed children.  DNA methylation signatures can function as biomarkers of prenatal smoke exposure. However, the robustness of these DNA methylation signatures across child ages, genetic ancestry groups, or tissues is not clear. Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study. In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was consistent in European and African genetic ancestry samples. In the full sample at age 15 (n=`r table(modeldata$childteen)[2]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score, and the association was attenuated among the European and Admixed - Latin genetic ancestry samples. The polymethylation score classified prenatal maternal smoke exposure accurately (AUC age 9=0.77, age 15=0.77). Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P*-value=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`). Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than an individual DNA methylation site. DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

>>>>

Prenatal maternal smoking is associated with low birthweight, neurological disorders, and asthma in exposed children.  DNA methylation signatures can function as biomarkers of prenatal smoke exposure. However, the robustness of these DNA methylation signatures across child ages, genetic ancestry groups, or tissues is not clear. Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study. In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was consistent in European and African genetic ancestry samples. In the full sample at age 15 (n=`r table(modeldata$childteen)[2]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score, and the association was attenuated among the European and Admixed - Latin genetic ancestry samples. The polymethylation score classified prenatal maternal smoke exposure accurately (AUC age 9=0.77, age 15=0.77). Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(5)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P*-value=`r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`). Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than an individual DNA methylation site. DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2022-05-05 Papers/Polymethylation_scores/05_results.Rmd
revisions may5

 All of the polymethylation scores for prenatal smoke exposure were correlated with each other (Supplemental Figure 1). Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P* value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).

Among the `r n_twovisit` individuals with data from both the ages 9 and 15 visits, we compared epigenetic measures over time (Supplemental Figure 2). The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the 2016 Joubert meta-analysis coefficients from newborn cord blood (Pearson $r$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P*-value\<0.0001). Global DNA methylation ($r$=`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P*-value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: $r$=`r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P*-value\<0.0001, GRIM: $r$=`r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P*-value\<0.0001) Epigenetic ages were higher at the age 15 visit than the age 9 visit, as expected (Supplemental Figure 3).The distributions of global methylation and polymethylation scores were more consistent (Supplemental Figure 3). Almost all saliva samples consisted of primarily immune cells, based on estimated cell-type proportions. At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`% (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`). At age 15, the mean estimated immune cell proportion was `r paste(round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2), '%')` (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25), na.rm=T), 2)*100`, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.75), na.rm=T), 2)*100`, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`).

@@

At age 9, prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting for base model covariates. Similarly, at age 15 prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)). Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3). In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`). In the Admixed Latin genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3).

>>>>

 All of the polymethylation scores for prenatal smoke exposure were correlated with each other (Supplemental Figure 1). Our primary polymethylation score for prenatal smoke exposure, which was calculated using coefficients from a cell-type corrected regression, was weakly correlated with estimated cell-type proportion of immune cells (Pearson $r$=`r round(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$estimate, 2)`; *P*-value `r my_p_val(cor.test(modeldata$SSnewbornCT_center, modeldata$Leukocytes_saliva)$p.value)`).

Among the `r n_twovisit` individuals with data from both the ages 9 and 15 visits, we compared epigenetic measures over time (Supplemental Figure 2). The correlation across ages was strongest for the polymethylation scores for prenatal smoke exposure constructed using the 2016 Joubert meta-analysis coefficients from newborn cord blood (Pearson $r$=`r round(cor(model_twovisit$SSnewbornCT_center_Age_9, model_twovisit$SSnewbornCT_center_Age_15), 2)`, *P*-value\<0.0001). Global DNA methylation ($r$=`r round(cor(model_twovisit$globalmethylation_Age_9, model_twovisit$globalmethylation_Age_15), 2)`, *P*-value\<0.0001) and epigenetic clocks were less strongly correlated across ages (Pediatric: $r$=`r round(cor(model_twovisit$PedBE_Age_9, model_twovisit$PedBE_Age_15), 2)`; *P*-value\<0.0001, GRIM: $r$=`r round(cor(model_twovisit$DNAmGrimAge_Age_9, model_twovisit$DNAmGrimAge_Age_15), 2)`; *P*-value\<0.0001) Epigenetic ages were higher at the age 15 visit than the age 9 visit, as expected (Supplemental Figure 3).The distributions of global methylation and polymethylation scores were more consistent (Supplemental Figure 3). Almost all saliva samples consisted of primarily immune cells, based on estimated cell-type proportions. At age 9, the mean estimated immune cell proportion was `r round(mean(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`% (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.25), na.rm=T), 2)*100`%, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_9, c(0.75), na.rm=T), 2)*100`%, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_9, na.rm=T), 2)*100`%). At age 15, the mean estimated immune cell proportion was `r paste(round(mean(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100, '%')` (minimum=`r round(min(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%, first quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.25), na.rm=T), 2)*100`%, third quartile=`r round(quantile(modeldata_wide$Leukocytes_saliva_Age_15, c(0.75), na.rm=T), 2)*100`%, maximum=`r round(max(modeldata_wide$Leukocytes_saliva_Age_15, na.rm=T), 2)*100`%).

@@

At age 9, prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95% CI: `r paste(round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`) after adjusting for base model covariates. Similarly, at age 15 prenatal maternal smoking was associated with `r -1*round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(estimate), 2)` percent lower DNA methylation at cg055975921 (95%CI: (`r paste(round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.low), 2), round(global_models %>% filter(childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921')%>%unnest(tidied)%>%filter(term=='smkPreg_binaryYes')%>%pull(conf.high), 2), sep=', ')`)). Prenatal maternal smoking remained significantly associated with a decrease in cg05575921 DNA methylation in the African genetic ancestry sample at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='African ancestry'", 'smkPreg_binaryYes', 2)`; Figure 3). In the European genetic ancestry sample, the association was consistent in direction at age 9 (`r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`) and age 15 (`r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='European ancestry'", 'smkPreg_binaryYes', 2)`). In the Admixed-Latin genetic ancestry sample, the association was no longer significant and not consistent in direction (Age 9: `r coefCI(local_models, "childteen=='Age 9' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`; Age 15: `r coefCI(local_models, "childteen=='Age 15' & predictors=='base model' & outcome=='AHRR: cg05575921' & ancestry=='Admixed ancestry - Latin heritage'", 'smkPreg_binaryYes', 2)`) (Figure 3).

### 2022-05-05 Papers/Polymethylation_scores/06_discussion.Rmd
revisions may5

In the longitudinal Fragile Families and Child Wellbeing birth cohort, prenatal maternal smoking was associated with saliva DNA methylation in children at ages 9 and 15. Children with prenatal smoke exposure had `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation scores at age 9 than children not exposed. These findings were robust across strata of genetic ancestry and children's age. These signatures were specific, as global DNA methylation and epigenetic clocks were not associated with prenatal smoke exposure. When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self-reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9. A single *a priori* DNA methylation site, *AHRR*:cg05575921, did not ( `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`percentage point increase). Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking. A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible. Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.

>>>>

In the longitudinal Fragile Families and Child Wellbeing birth cohort, prenatal maternal smoking was associated with saliva DNA methylation in children at ages 9 and 15. Children with prenatal smoke exposure had `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation scores at age 9 than children not exposed. These findings were robust across strata of genetic ancestry and children's age. These signatures were specific, as global DNA methylation and epigenetic clocks were not associated with prenatal smoke exposure. When assessing biomarker utility, the polymethylation score for prenatal smoke exposure improved the classification of self-reported prenatal maternal smoking over a base model by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'polymethylation score'))%>%pull(5)` percentage points at age 9. A single *a priori* DNA methylation site, *AHRR*: cg05575921, did not ( `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg05575921'))%>%pull(5)`percentage point increase). Our results suggest salivary DNA methylation can be used as an alternative to peripheral blood DNA methylation for biomarkers of prenatal maternal smoking. A saliva-based DNA methylation biomarker may be particularly useful in cases where blood collection, processing, and storage is not possible. Saliva can often be collected directly by participants and saliva sample collection kits have stabilizers allowing room temperature storage prior to DNA methylation measurement.

### 2022-05-05 Papers/Polymethylation_scores/99_FigureCaptions.Rmd
revisions may5

#### Table 1: Bivariate associations between prenatal maternal smoking (no/yes) and selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study

```{r,  eval=print_figs, include=T}
smoking_tbl
```

@@

#### **Figure 5 - Comparison of different polymethylation scores and their accuracy at classifying prenatal smoke exposure among `r n_total` children with saliva samples in the Fragile Families and Child Wellbeing study as opposed to previously reported results in cord and peripheral blood samples

```{r, echo=F, eval=print_figs, include=T,  fig.width=12}
auc_compare
```

>>>>

#### **Table 1: Bivariate associations between prenatal maternal smoking (no/yes) and selected DNA methylation summary measures and important covariates among a diverse sample of `r n_total` children in the Fragile Families and Child Wellbeing study**

```{r, echo=F, eval=print_figs, include=T, include=T}
smoking_tbl
```

@@

#### **Figure 5 - Comparison of different polymethylation scores and their accuracy at classifying prenatal smoke exposure among `r n_total` children with saliva samples in the Fragile Families and Child Wellbeing study as opposed to previously reported results in cord and peripheral blood samples**

```{r, echo=F, eval=print_figs, warning=F, include=T,  fig.width=12}
auc_compare
```
